\textit{Supplementary materials for}: Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-15 paper/main.tex
another pass through main text + supplement, recompiled changes doc
-- sending this version to co-authors for comments...

Following our prior work~\citep{HeusEtal17}, we used a permutation-based
correction procedure to help isolate the behavioral aspects of clustering that
we were most interested in. After computing the uncorrected clustering score
(for the given list and observed recall sequence), we compute a ``null''
distribution of $n$ additional clustering scores after randomly shuffling the
order of the recalled words (we use $n = 500$ in the present study). This null
distribution represents an approximation of the range of clustering scores one
might expect to observe by ``chance,'' given that a hypothetical participant
was \textit{not} truly clustering their recalls, but where the hypothetical
participant still studied and recalled exactly the same items (with the same
features) as the true participant. We define the \textit{permutation-corrected
clustering score} as the percentile rank of the observed uncorrected clustering
score in this estimated null distribution. In this way, a corrected score of 1
indicates that the observed score was greater than any clustering score one
might expect by chance---in other words, good evidence that the participant was
truly clustering their recalls along the given feature dimension. We applied
this correction procedure to all of the clustering scores (feature and
temporal) reported in this paper.

@@

We developed a permutation-based procedure to factor out the effects of
temporal clustering from the clustering scores for each feature dimension. For
a given set of recalled items (whose presentation positions are given by $x_1,
x_2, x_3, …, x_N$), we circularly shift the presentation positions by a
randomly chosen amount (between 1 and the list length) to obtain a new set of
items. Since the new set of items will have the same (average) temporal
distances between successive recalls, the temporal clustering score for the new
set of items is equal (on average) to the temporal clustering score for the
original recalls. However, we can then re-compute the feature clustering score
for those new items. Finally, we can compute a ``temporally corrected'' feature
clustering score by computing the average percentile rank of the observed (raw)
feature clustering score within the distributions of circularly shifted feature
clustering scores, across $N = 500$ repetitions of this procedure. This new
temporally corrected score provides an estimate of the observed degree of
feature clustering over and above what could be accounted for by temporal
clustering alone.

@@

A core assumption of our approach is that each participant organizes
their memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order in their recall sequences (and,
presumably, organize in memory) the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints computed
for different participants. We reasoned that the feature rich condition would
provide the best opportunity to test this assumption, since the clustering
scores would not be potentially confounded by order manipulations. To test our
``unique memory fingerprint'' assumption, we compared the similarity
(correlation) between the fingerprint from a single list (from one participant)
and (a) the average fingerprint from all other lists from the same participant
versus (b) the average fingerprint from each other participant (across all of
their lists). We found that participants' fingerprints on a held-out list are
reliably more similar to the same participant's fingerprints on other lists
than to other participants' fingerprints ($t(70280) = 5.077, p < 0.001, d =
0.162, \mathrm{CI} = [3.086, 6.895]$). This suggests that participants'
fingerprints are stable across lists, and that each participant's fingerprint
is unique to them.

We also wondered whether adding these incidental visual features to later lists
(after the participants had already studied impoverished lists), or removing
the visual features from later lists (after the participants had already
studied visually diverse lists) might affect memory performance. In other
words, we sought to test for potential effects of changing the ``richness'' of
participants' experiences over time. All participants studied and recalled a
total of 16 lists; we defined \textit{early} lists as the first eight lists and
\textit{late} lists as the last eight lists each participant encountered. To
help interpret our results, we compared participants' memories on early versus
late lists in the above feature rich and reduced conditions. Participants in
both conditions remembered more words on early versus late lists (feature rich:
$t(66) = 4.553, p < 0.001, d = 0.233, \mathrm{CI} = [2.427, 7.262]$; reduced:
$t(60) = 2.434, p = 0.018, d = 0.134, \mathrm{CI} = [0.493, 4.910]$).
Participants in the feature rich (but not reduced) conditions exhibited more
temporal clustering on early versus late lists (feature rich: $t(66) = 2.268, p
= 0.027, d = 0.181, \mathrm{CI} = [0.437, 4.425]$; reduced: $t(60) = 0.986, p =
0.328, d = 0.061, \mathrm{CI} = [-0.897, 3.348]$). And participants in both
conditions tended to exhibit more semantic clustering on early versus late
lists (feature rich, category: $t(66) = 3.684, p < 0.001, d = 0.220,
\mathrm{CI} = [1.733, 5.732]$; feature rich, size: $t(66) = 1.629, p = 0.108, d
= 0.100, \mathrm{CI} = [-0.207, 3.905] $; reduced, category: $t(60) = 2.755, p
= 0.008, d = 0.177, \mathrm{CI} = [0.761, 5.189]$; reduced, size: $t(60) =
3.081, p = 0.003, d = 0.201, \mathrm{CI} = [1.210, 5.326]$). Participants in
the reduced (but not feature rich) conditions tended to exhibit more
lexicographic clustering on early versus late lists (feature rich, word length:
$t(66) = -0.100, p = 0.921, d = -0.010, \mathrm{CI} = [-2.217, 1.899]$; feature
rich, first letter: $t(66) = -0.412, p = 0.681, d = -0.045, \mathrm{CI} =
[-2.461, 1.645]$; reduced, word length: $t(60) = 3.762, p < 0.001, d = 0.261,
\mathrm{CI} = [1.604, 6.821]$; reduced, first letter: $t(60) = 1.721, p =
0.090, d = 0.175, \mathrm{CI} = [-0.138, 4.098]$). Taken together, these
comparisons suggest that even when the presence or absence of incidental visual
features is stable across lists, participants still exhibit some differences in
their performance and memory organization tendencies for early versus late
lists.

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to exhibit stronger clustering effects on feature rich
(versus reduced) lists, and (b) participants tended to remember more words and
exhibit stronger clustering effects on early (versus late) lists, we expected
these early versus late differences to be enhanced in the reduced (early)
condition and diminished in the reduced (late) condition. However, to our
surprise, participants in \textit{neither} condition exhibited reliable early
versus late differences in accuracy (reduced (early): $t(41) = 1.499, p =
0.141, d = 0.098, \mathrm{CI} = [-0.345, 3.579]$; reduced (late): $t(40) =
1.462, p = 0.152, d = 0.121, \mathrm{CI} = [-0.376, 2.993]$), temporal
clustering (reduced (early): $t(41) = 0.857, p = 0.396, d = 0.068, \mathrm{CI}
= [-1.012, 2.896]$; reduced (late): $t(40) = 1.244, p = 0.221, d = 0.128,
\mathrm{CI} = [-0.894, 3.088]$), nor feature-based clustering (reduced (early),
category: $t(41) = 0.707, p = 0.484, d = 0.068, \mathrm{CI} = [-1.314, 2.830]$;
reduced (early), size: $t(41) = 0.803, p = 0.427, d = 0.079, \mathrm{CI} =
[-1.142, 2.953]$; reduced (early), length: $t(41) = 0.461, p = 0.648, d =
0.060, \mathrm{CI} = [-1.545, 2.462]$; reduced (early), first letter: $t(41) =
0.781, p = 0.439, d = 0.101, \mathrm{CI} = [-1.039, 2.881]$; reduced (late),
category: $t(40) = -0.101, p = 0.920, d = -0.009, \mathrm{CI} = [-2.307,
1.776]$; reduced (late), size: $t(40) = 0.555, p = 0.582, d = 0.058,
\mathrm{CI} = [-1.444, 2.274] $; reduced (late), length: $t(40) = 1.482, p =
0.146, d = 0.126, \mathrm{CI} = [-0.444, 3.743]$; reduced (late), first letter:
$t(40) = -0.143, p = 0.887, d = -0.017, \mathrm{CI} = [-2.204, 1.830] $). We
hypothesized that adding or removing the variability in the visual features was
acting as a sort of ``event boundary'' between early and late
lists~\citep[e.g., ][]{ClewEtal19,RadvZack17,RadvCope06}. In prior work, we
(and others) have found that memories formed just after event boundaries can be
enhanced~\citep[e.g., due to less contextual interference between pre- and
post-boundary items;][]{MannEtal16, PettEtal16, GoldEtal17, FlorEtal17}.

@@

indistinguishable from performance on the early feature rich lists ($\|t\|$s

@@

When we closely examined the sequences of words participants recalled from
early order-manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One difference is
that participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
increased tendency to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncrasies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing incidental visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

@@

At the group level, there seemed to be almost no lingering impact of sorting
early lists on memory for later lists. To simplify the presentation, we report
these null results in aggregate across the three feature groupings. Relative to
memory performance on late feature rich lists, participants' memory performance
in all six order manipulation conditions showed no reliable differences
(semantic: $t(125) = 0.487, p = 0.627, d = 0.087, \mathrm{CI} = [-1.661,
2.323]$; lexicographic: $t(125) = 0.878, p = 0.382, d = 0.156, \mathrm{CI} =
[-1.226, 3.044]$; visual: $t(126) = 1.437, p = 0.153, d = 0.254, \mathrm{CI} =
[-0.447, 3.519]$). Nor did we observe any reliable differences in temporal
clustering on late lists (relative to late feature rich lists; semantic:
$t(125) = 0.157, p = 0.875, d = 0.028, \mathrm{CI} = [-1.859, 1.974]$;
lexicographic: $t(125) = 0.998, p = 0.320, d = 0.177, \mathrm{CI} = [-0.902,
2.920]$; visual: $t(126) = 0.548, p = 0.585, d = 0.097, \mathrm{CI} = [-1.450,
2.365]$). Aside from a slightly increased tendency for participants to cluster
words by their length on late visual order manipulation lists (more than late
feature rich lists; $t(126) = 2.005, p = 0.047, d = 0.355, \mathrm{CI} =
[0.211, 3.722]$), we observed no reliable differences in any type of feature
clustering on late order manipulation condition lists versus late feature rich
lists ($\|t\|$s $\leq 1.124, p$s $\geq 0.263$).

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. As
described above, a participant's memory fingerprint describes how they tend to
retrieve memories of the studied items, perhaps searching in parallel through
several feature spaces (or along several representational dimensions). To gain
insights into the dynamics of how participants' clustering scores tended to
change over time, we computed the average (across participants) fingerprint
from each list, from each order manipulation condition
(Fig.~\ref{fig:fingerprint-trajectories}). We projected these fingerprints into
a two-dimensional space to help visualize the dynamics (top panels; see
\textit{Computing low-dimensional embeddings of memory fingerprints}). We found
that participants' average fingerprints tended to remain relatively stable on
early lists, and exhibited a ``jump'' to another stable state on later lists.
The sizes of these jumps varied somewhat across conditions (the Euclidean
distances between fingerprints in their original high dimensional spaces are
displayed in the bottom panels). We also averaged the fingerprints across early
and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations
for those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

    \caption{\textbf{Memory fingerprint dynamics (order manipulation
    conditions).} \textbf{A.} Each column (and color) reflects an experimental
    condition. In the top panels, each marker displays a 2D projection of the
    (across-participant) average memory fingerprint for one list. Order
    manipulation (early) lists are denoted by circles and randomly ordered
    (late) lists are denoted by stars. All of the fingerprints (across all
    conditions and lists) are projected into a common space. The bar plots in
    the bottom panels display the Euclidean distances of the per-list memory
    fingerprints to the list 0 fingerprint, for each condition. Error bars
    denote bootstrap-estimated 95\% confidence intervals. The dotted vertical
    lines denote the boundaries between early and late lists. \textbf{B.} In
    this panel, the fingerprints for early (circle) and late (star) lists are
    averaged across lists and participants before projecting the fingerprints
    into a (new) 2D space. See Figure~\fingerprintTrajectoryRandom~for
    analogous plots for the random conditions. }
    \label{fig:fingerprint-trajectories}

@@

feature clustering on early lists tended to recall more items on late lists
(across conditions: $r(179) = 0.492, p < 0.001$; all conditions individually:
$r$s $\geq 0.462$, all $p$s $\leq 0.010$). Participants who recalled more items
on early lists also tended to show stronger feature clustering on late lists
(across conditions: $r(179) = 0.280, p < 0.001$; all non-visual conditions:
$r$s $\geq 0.445$, all $p$s $\leq 0.014$; color: $r(29) = 0.298, p = 0.103$;
location: $r(28) = 0.354, p = 0.055$). Neither of these effects showed
condition-level trends (early feature clustering versus late recall
probability: $r(4) = -0.299, p = 0.565$; early recall probability versus late
feature clustering: $r(4) = 0.400, p = 0.432$). We also looked for associations
between feature clustering and temporal clustering. Across every order
manipulation condition, participants who exhibited stronger feature clustering
also exhibited stronger temporal clustering. For early lists

@@

0.924, p < 0.001$), for each condition individually (all $r$s $\geq 0.822$, all
$p$s $< 0.001$), and across conditions ($r(4) = 0.964, p = 0.002$). For late
lists (Fig.~\ref{fig:clustering-scatterplots}F), the results were more variable

@@

feature clustering on early lists trended towards showing stronger temporal

>>>>

Following our prior work~\citep{HeusEtal17}, we used a permutation-based
correction procedure to help isolate the behavioral aspects of clustering that
we were most interested in. After computing the uncorrected clustering score
(for the given list and observed recall sequence), we compute a ``null''
distribution of $n$ additional clustering scores after randomly shuffling the
order of the recalled words (we use $n = 500$ in the present study). This null
distribution represents an approximation of the range of clustering scores one
might expect to observe by ``chance,'' given that a hypothetical participant
was \textit{not} truly clustering their recalls, but where the hypothetical
participant still studied and recalled exactly the same items (with the same
features) as the true participant. We define the \textit{permutation-corrected
clustering score} as the percentile rank of the observed uncorrected clustering
score in this estimated null distribution. In this way, a corrected score of 1
indicates that the observed score was greater than any clustering score one
might expect by chance---in other words, good evidence that the participant was
truly clustering their recalls along the given feature dimension. We applied
this correction procedure to all of the clustering scores (feature and
temporal) reported in this paper. In Figure~\clusterCorrs~we report how
participants' clustering scores along different feature dimensions (in the
order manipulation conditions) are correlated, and how clustering scores change
across lists.

@@

We developed a permutation-based procedure to factor out the effects of
temporal clustering from the clustering scores for each feature dimension. For
a given set of recalled items (whose presentation positions are given by $x_1,
x_2, x_3, …, x_N$), we circularly shift the presentation positions by a
randomly chosen amount (between 1 and the list length) to obtain a new set of
items at the (now altered) positions of the original recalls. Since the new
set of items will have the same (average) temporal distances between successive
recalls, the temporal clustering score for the new set of items is equal (on
average) to the temporal clustering score for the original recalls. However, we
can then re-compute the feature clustering score for those new items. Finally,
we can compute a ``temporally corrected'' feature clustering score by computing
the average percentile rank of the observed (raw) feature clustering score
within the distributions of circularly shifted feature clustering scores,
across $N = 500$ repetitions of this procedure. This new temporally corrected
score provides an estimate of the observed degree of feature clustering over
and above what could be accounted for by temporal clustering alone.

@@

A core assumption of our approach is that each participant organizes their
memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order, in their recall sequences (and,
presumably, organize in memory), the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints
computed for different participants. We reasoned that the feature rich
condition would provide the best opportunity to test this assumption, since the
clustering scores would not be potentially confounded by order manipulations.
To test our ``unique memory fingerprint'' assumption, we compared the
similarity (correlation) between the fingerprint from a single list (from one
participant) and (a) the average fingerprint from all other lists from the same
participant versus (b) the average fingerprint from each other participant
(across all of their lists). Repeating this procedure for all lists and
participants, we found that participants' fingerprints on a held-out list are
reliably more similar to the same participant's fingerprints on other lists
than to other participants' fingerprints ($t(70280) = 5.077, p < 0.001, d =
0.162, \mathrm{CI} = [3.086, 6.895]$). This suggests that participants'
fingerprints are stable across lists, and that each participant's fingerprint
is unique to them.

We next asked whether adding these incidental visual features to later lists
(after the participants had already studied impoverished lists), or removing
the visual features from later lists (after the participants had already
studied visually diverse lists) might affect memory performance. In other
words, we sought to test for potential effects of changing the ``richness'' of
participants' experiences over time. All participants studied and recalled a
total of 16 lists; we defined \textit{early} lists as the first eight lists and
\textit{late} lists as the last eight lists each participant encountered. To
help interpret our results, we compared participants' memories on early versus
late lists in the above feature rich and reduced conditions. Participants in
both conditions remembered more words on early versus late lists (feature rich:
$t(66) = 4.553, p < 0.001, d = 0.233, \mathrm{CI} = [2.427, 7.262]$; reduced:
$t(60) = 2.434, p = 0.018, d = 0.134, \mathrm{CI} = [0.493, 4.910]$).
Participants in the feature rich (but not reduced) conditions exhibited more
temporal clustering on early versus late lists (feature rich: $t(66) = 2.268, p
= 0.027, d = 0.181, \mathrm{CI} = [0.437, 4.425]$; reduced: $t(60) = 0.986, p =
0.328, d = 0.061, \mathrm{CI} = [-0.897, 3.348]$). And participants in both
conditions tended to exhibit more semantic clustering on early versus late
lists (feature rich, category: $t(66) = 3.684, p < 0.001, d = 0.220,
\mathrm{CI} = [1.733, 5.732]$; feature rich, size: $t(66) = 1.629, p = 0.108, d
= 0.100, \mathrm{CI} = [-0.207, 3.905] $; reduced, category: $t(60) = 2.755, p
= 0.008, d = 0.177, \mathrm{CI} = [0.761, 5.189]$; reduced, size: $t(60) =
3.081, p = 0.003, d = 0.201, \mathrm{CI} = [1.210, 5.326]$). Participants in
the reduced (but not feature rich) conditions tended to exhibit more
lexicographic clustering on early versus late lists (feature rich, word length:
$t(66) = -0.100, p = 0.921, d = -0.010, \mathrm{CI} = [-2.217, 1.899]$; feature
rich, first letter: $t(66) = 0.412, p = 0.681, d = 0.045, \mathrm{CI} =
[-1.645, 2.461]$; reduced, word length: $t(60) = 3.762, p < 0.001, d = 0.261,
\mathrm{CI} = [1.604, 6.821]$; reduced, first letter: $t(60) = 1.721, p =
0.090, d = 0.175, \mathrm{CI} = [-0.138, 4.098]$). Taken together, these
comparisons suggest that even when the presence or absence of incidental visual
features is stable across lists, participants still exhibit some differences in
their performance and memory organization tendencies for early versus late
lists.

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to exhibit stronger clustering effects on feature rich
(versus reduced) lists, and (b) participants tended to remember more words and
exhibit stronger clustering effects on early (versus late) lists, we expected
these early versus late differences to be enhanced in the reduced (early)
condition and diminished in the reduced (late) condition. However, to our
surprise, participants in \textit{neither} condition exhibited reliable early
versus late differences in accuracy (reduced (early): $t(41) = 1.499, p =
0.141, d = 0.098, \mathrm{CI} = [-0.345, 3.579]$; reduced (late): $t(40) =
1.462, p = 0.152, d = 0.121, \mathrm{CI} = [-0.376, 2.993]$), temporal
clustering (reduced (early): $t(41) = 0.857, p = 0.396, d = 0.068, \mathrm{CI}
= [-1.012, 2.896]$; reduced (late): $t(40) = 1.244, p = 0.221, d = 0.128,
\mathrm{CI} = [-0.894, 3.088]$), nor feature-based clustering (reduced (early),
category: $t(41) = 0.707, p = 0.484, d = 0.068, \mathrm{CI} = [-1.314, 2.830]$;
reduced (early), size: $t(41) = 0.803, p = 0.427, d = 0.079, \mathrm{CI} =
[-1.142, 2.953]$; reduced (early), length: $t(41) = 0.461, p = 0.648, d =
0.060, \mathrm{CI} = [-1.545, 2.462]$; reduced (early), first letter: $t(41) =
0.781, p = 0.439, d = 0.101, \mathrm{CI} = [-1.039, 2.881]$; reduced (late),
category: $t(40) = 0.101, p = 0.920, d = 0.009, \mathrm{CI} = [-1.776, 2.307]$;
reduced (late), size: $t(40) = 0.555, p = 0.582, d = 0.058, \mathrm{CI} =
[-1.444, 2.274] $; reduced (late), length: $t(40) = 1.482, p = 0.146, d =
0.126, \mathrm{CI} = [-0.444, 3.743]$; reduced (late), first letter: $t(40) =
-0.143, p = 0.887, d = -0.017, \mathrm{CI} = [-2.204, 1.830] $). We
hypothesized that adding or removing the variability in the visual features was
acting as a sort of ``event boundary'' between early and late
lists~\citep[e.g., ][]{ClewEtal19,RadvZack17,RadvCope06}. In prior work, we
(and others) have found that memories formed just after event boundaries can be
enhanced~\citep[e.g., due to less contextual interference between pre- and
post-boundary items;][]{MannEtal16, PettEtal16, GoldEtal17, FlorEtal17}.

@@

indistinguishable from performance on the early feature rich lists ($|t|$s

@@

When we closely examined the sequences of words participants recalled from
early order-manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One difference is
that participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}C; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Figs.~\ref{fig:recall-dynamics}B, C), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Figs.~\ref{fig:recall-dynamics}B, C), and participants also seemed to display an
increased tendency to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncrasies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing incidental visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

@@

At the group level, there seemed to be almost no lingering impact of sorting
early lists on memory for later lists. To simplify the presentation, we report
these null results in aggregate across the three feature groupings. Relative to
memory performance on late feature rich lists, participants' memory performance
in all six order manipulation conditions showed no reliable differences
(semantic: $t(125) = 0.487, p = 0.627, d = 0.087, \mathrm{CI} = [-1.661,
2.323]$; lexicographic: $t(125) = 0.878, p = 0.382, d = 0.156, \mathrm{CI} =
[-1.226, 3.044]$; visual: $t(126) = 1.437, p = 0.153, d = 0.254, \mathrm{CI} =
[-0.447, 3.519]$). Nor did we observe any reliable differences in temporal
clustering on late lists (relative to late feature rich lists; semantic:
$t(125) = 0.157, p = 0.875, d = 0.028, \mathrm{CI} = [-1.859, 1.974]$;
lexicographic: $t(125) = 0.998, p = 0.320, d = 0.177, \mathrm{CI} = [-0.902,
2.920]$; visual: $t(126) = 0.548, p = 0.585, d = 0.097, \mathrm{CI} = [-1.450,
2.365]$). Aside from a slightly increased tendency for participants to cluster
words by their length on late visual order manipulation lists (more than late
feature rich lists; $t(126) = 2.005, p = 0.047, d = 0.355, \mathrm{CI} =
[0.211, 3.722]$), we observed no reliable differences in any type of feature
clustering on late order manipulation condition lists versus late feature rich
lists ($|t|$s $\leq 1.124, p$s $\geq 0.263$).

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. As
described above, a participant's memory fingerprint characterizes how they tend
to retrieve memories of the studied items, perhaps searching in parallel
through several feature spaces (or along several representational dimensions).
To gain insights into the dynamics of how participants' clustering scores
tended to change over time, we computed the average (across participants)
fingerprint from each list, from each order manipulation condition
(Fig.~\ref{fig:fingerprint-trajectories}). We projected these fingerprints into
a two-dimensional space to help visualize the dynamics (top panels; see
\textit{Computing low-dimensional embeddings of memory fingerprints}). We found
that participants' average fingerprints tended to remain relatively stable on
early lists, and exhibited a ``jump'' to another stable state on later lists.
The sizes of these jumps varied somewhat across conditions (the Euclidean
distances between fingerprints in their original high dimensional spaces are
displayed in the bottom panels). We also averaged the fingerprints across early
and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations
for those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

    \caption{\textbf{Memory fingerprint dynamics (order manipulation
    conditions).} \textbf{A.} Each column (and color) reflects an experimental
    condition. In the top panels, each marker displays a 2D projection of the
    (across-participant) average memory fingerprint for one list. Order
    manipulation (early) lists are denoted by circles and randomly ordered
    (late) lists are denoted by stars. All of the fingerprints (across all
    conditions and lists) are projected into a common space. The bar plots in
    the bottom panels display the Euclidean distances of the per-list memory
    fingerprints to the average fingerprint across all prior lists, for each
    condition. Error bars denote bootstrap-estimated 95\% confidence intervals.
    The dotted vertical lines denote the boundaries between early and late
    lists. \textbf{B.} In this panel, the fingerprints for early (circle) and
    late (star) lists are averaged across lists and participants before
    projecting the fingerprints into a (new) 2D space. See
    Figure~\fingerprintTrajectoryRandom~for analogous plots for the random
    conditions. } \label{fig:fingerprint-trajectories}

@@

feature clustering on early lists in the non-visual conditions tended to recall
more items on late lists (across conditions: $r(179) = 0.230, p = 0.002,
\mathrm{CI} = [0.072, 0.372]$; all non-visual conditions individually: $r$s
$\geq 0.405$, all $p$s $\leq 0.027$; color: $r(29) = 0.212, p = 0.251,
\mathrm{CI} = [-0.164, 0.532]$; location: $r(28) = 0.320, p = 0.085,
\mathrm{CI} = [0.011, 0.584]$). Participants who recalled more items on early
lists also tended to show stronger feature clustering on late lists (across
conditions: $r(179) = 0.464, p < 0.001, \mathrm{CI} = [0.321, 0.582]$;
individual conditions: all $r$s $\geq 0.377$, all $p$s $\leq 0.040$). Neither
of these effects showed condition-level trends (early feature clustering versus
late recall probability: $r(4) = -0.338, p = 0.512, \mathrm{CI} = [-0.971,
0.634]$; early recall probability versus late feature clustering: $r(4) =
0.451, p = 0.369, \mathrm{CI} = [-0.986, 0.998]$). We also looked for
associations between feature clustering and temporal clustering. Across every
order manipulation condition, participants who exhibited stronger feature
clustering also exhibited stronger temporal clustering. For early lists

@@

0.916, p < 0.001, \mathrm{CI} = [0.893, 0.936]$), for each condition
individually (all $r$s $\geq 0.822$, all $p$s $< 0.001$), and across conditions
($r(4) = 0.964, p = 0.002$). For late lists
(Fig.~\ref{fig:clustering-scatterplots}F), the results were more variable

@@

feature clustering on early lists showed stronger temporal

### 2023-11-15 paper/supplement.tex
another pass through main text + supplement, recompiled changes doc
-- sending this version to co-authors for comments...

    \caption{\textbf{Memory fingerprint dynamics (random conditions).}
    \textbf{A.} Each column (and color) reflects an experimental condition. In
    the top panels, each marker displays a 2D projection of the
    (across-participant) average memory fingerprint for one list. Early lists
    are denoted by circles and late lists are denoted by stars. All of the
    fingerprints (across all conditions and lists) are projected into a common
    space. The bar plots in the bottom panels display the Euclidean distances
    of the per-list memory fingerprints to the list 0 fingerprint, for each
    condition. Error bars denote bootstrap-estimated 95\% confidence intervals.
    The dotted vertical lines denote the boundaries between early and late
    lists. \textbf{B.} In this panel, the fingerprints for early (circle) and
    late (star) lists are averaged across lists and participants before
    projecting the fingerprints into a (new) 2D space.} \label{fig:fingerprint-trajectories-random}

>>>>

    \caption{\textbf{Memory fingerprint dynamics (random conditions).}
    \textbf{A.} Each column (and color) reflects an experimental condition. In
    the top panels, each marker displays a 2D projection of the
    (across-participant) average memory fingerprint for one list. Early lists
    are denoted by circles and late lists are denoted by stars. All of the
    fingerprints (across all conditions and lists) are projected into a common
    space. The bar plots in the bottom panels display the Euclidean distances
    of the per-list memory fingerprints to the average fingerprint across all
    prior lists, for each condition. Error bars denote bootstrap-estimated 95\%
    confidence intervals. The dotted vertical lines denote the boundaries
    between early and late lists. \textbf{B.} In this panel, the fingerprints
    for early (circle) and late (star) lists are averaged across lists and
    participants before projecting the fingerprints into a (new) 2D space.}
    \label{fig:fingerprint-trajectories-random}
